Normalization: A pipeline to analyse time-course gene expression data

### 2019-05-02 scripts/00_introduction.Rmd
Adding schemas of the workflow + rewording

While there are many methods that have been proposed for discrete aspects of
time course data, the entire workflow for analysis of such data remains
difficult, particularly for long, developmental time series. Most methods
proposed for time course data are concerned with detecting genes that are
changing over time (differential expression analysis), examples being
`edge` [@storey:significance], functional component analysis based models
[@wu:more], time-course permutation tests [@park:statistical], XXX
[@sun:multiple]
**TODO: Other methods to mention: impulse models, what else?
Citations!**. However, with long time course datasets, particularly in
developmental systems, a massive number of genes will show *some* change.
For example, in the \xxx system, over \xxx\% of genes are shown to be changing
over time. The task in these settings is often not to detect changes in genes,
but to categorize into biologically interpretable patterns the vast number of
changes discovered.

@@

This workflow represents an integration of both novel implementations of
previously established methods and new methodologies for the settings of
developmental time series. We provide the various steps of the workflow as
functions in a R package **FIXME**.

---
# Old intro

Time-course gene expression analysis workflows include steps used more broadly
for gene expression data analysis such as quality control (QC), exploratory
data analysis (EDA), pre-processing, normalization, and visualization. Yet, a
typical gene expression data analysis workflow is not entirely adequate for
analysing time-course data. In particular, differential expression analysis
needs to be adapted to time-course data, and often needs to be followed by a
clustering step to group genes sharing similar patterns of expression.

Here, we propose an integrated workflow for downstream analysis of time-course
data, with the following main steps: (1) quality control and normalization;
(2) differential expression analysis of time-course data; (3) clustering; (4)
downstream analysis of clusters with GO term, pathway, and motif enrichment.

This workflow relies on several standard packages for analysing gene
expression data, some specific for time-course data, others broadly used by
the community. We also provide alongside this workflow a package that nicely
wraps all calls to dependencies with a common API by defining a model object
holding all necessary metadata to perform all steps of the pipeline.
---

>>>>

While there are many methods that have been proposed for discrete aspects of
time course data, the entire workflow for analysis of such data remains
difficult, particularly for long, developmental time series. Most methods
proposed for time course data are concerned with detecting genes that are
changing over time (differential expression analysis), examples being `edge`
[@storey:significance], functional component analysis based models [@wu:more],
time-course permutation tests [@park:statistical], and multiple testing
strategies to combine singe time point differential expression analysis
[@sun:multiple]. However, with long time course datasets, particularly in
developmental systems, a massive number of genes will show *some* change. For
example, in the \xxx system, over \xxx\% of genes are shown to be changing
over time. The task in these settings is often not to detect changes in genes,
but to categorize into biologically interpretable patterns the vast number of
changes discovered.

@@

This workflow represents an integration of both novel implementations of
previously established methods and new methodologies for the settings of
developmental time series. It relies on several standard packages for
analysing gene expression data, some specific for time-course data, others
broadly used by the community. We provide the various steps of the workflow as
functions in a R package **FIXME**.

```{r schema, echo=FALSE, fig.width=8, fig.cap="Workflow for analyzing time-course datasets."}
library(knitr)
knitr::include_graphics("images/fig1_schemas.png")
```

### 2019-05-02 scripts/01_quality_control_normalization.Rmd
Adding schemas of the workflow + rewording

---
title: "Normalization: A pipeline to analyse time-course gene expression data"
output: rmarkdown::html_vignette
---

@@

 XXX The original paper says the data is "averaged over unique probes IDs."

@@

After filtering, we are left with `r n_probes` probes of interest in `r
n_samples`.

```{r}
# Now load in the metadata
# meta = read.delim("data/shoemaker2015/mice_meta.tsv", sep="\t")
data(shoemaker2015)
meta = shoemaker2015$meta
write.table(data, ".results/quantile_normalized.txt")
```

>>>>

---
title: "Normalization: A pipeline to analyse time-course gene expression data"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
---

@@

@@

After filtering, we are left with `r n_probes` probes of interest in 
`r n_samples` samples.

Now that we have normalized and filtered the gene expressions, let's load the
metadata. The package `moanin` contains the normalized data and meta of
[@shoemaker:ultrasensitive].

```{r}
# Now load in the metadata
data(shoemaker2015)
meta = shoemaker2015$meta
```

```{r echo=FALSE, results="hide"}
write.table(data, ".results/quantile_normalized.txt")
```
